---
title: "🌳 Tree Import (Simple)"
author: "Digital Forest Twin Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: bootstrap
---

# Overview

**Import prepared CSV data into the Digital Forest Twin database.**

This notebook is designed for **template-formatted CSVs** that have already been prepared using `prepare_ecosense.ipynb` or similar preprocessing.

## Prerequisites

Your CSV should have columns matching the template format:

| Column | Required | Description |
|--------|----------|-------------|
| `LocationID` | ✓ | FK to shared.Locations |
| `SpeciesID` | ✓ | FK to shared.Species |
| `Latitude` | ✓ | WGS84 latitude |
| `Longitude` | ✓ | WGS84 longitude |
| `DBH_cm` | | Diameter in centimeters |
| `Height_m` | | Height in meters |
| `ExternalID` | | Unique ID for linking |

See `data/templates/README.md` for full specification.

---

# Step 1: Setup & Database Connection

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r dependencies}
library(RPostgres)
library(DBI)
library(readr)
library(dplyr)
library(knitr)

# Load environment from docker/.env
readRenviron("../docker/.env")

cat("✓ Dependencies loaded\n")
```

```{r database-config}
# Database connection settings
postgres_host <- "localhost"
postgres_user <- Sys.getenv("POSTGRES_USER", "postgres")
postgres_password <- Sys.getenv("POSTGRES_PASSWORD")
postgres_db <- Sys.getenv("POSTGRES_DB", "postgres")
postgres_port <- as.integer(Sys.getenv("POSTGRES_PORT", "5432"))
pooler_tenant_id <- Sys.getenv("POOLER_TENANT_ID", "")

# Supavisor requires tenant ID in username
if (pooler_tenant_id != "") {
  postgres_user_pooler <- sprintf("%s.%s", postgres_user, pooler_tenant_id)
} else {
  postgres_user_pooler <- postgres_user
}

if (postgres_password == "") {
  stop("❌ POSTGRES_PASSWORD not found in docker/.env")
}

cat(sprintf("✓ Database: %s@%s:%d/%s\n", 
            postgres_user_pooler, postgres_host, postgres_port, postgres_db))
```

```{r database-connect}
# Create connection function
get_connection <- function() {
  dbConnect(
    RPostgres::Postgres(),
    host = postgres_host,
    user = postgres_user_pooler,
    password = postgres_password,
    dbname = postgres_db,
    port = postgres_port
  )
}

# Test connection
tryCatch({
  conn <- get_connection()
  dbDisconnect(conn)
  cat("✓ Database connection successful\n")
}, error = function(e) {
  cat(sprintf("❌ Connection failed: %s\n", e$message))
  cat("\nTroubleshooting:\n")
  cat("  1. Ensure Docker containers are running: docker compose up -d\n")
  cat("  2. Check docker/.env has correct POSTGRES_PASSWORD\n")
})
```

---

# Step 2: Load Prepared CSV

```{r config}
# ============================================================================
# CONFIGURATION - Set your CSV path here
# ============================================================================

CSV_PATH <- "../data/ecosense_250911_trees_prepared.csv"

# Alternative paths:
# CSV_PATH <- "../data/mathisle_250904_prepared.csv"
# CSV_PATH <- "../data/templates/trees_import_template.csv"

# ============================================================================
```

```{r load-csv}
if (!file.exists(CSV_PATH)) {
  cat(sprintf("❌ File not found: %s\n", CSV_PATH))
  cat("\nAvailable files in ../data/:\n")
  for (f in list.files("../data", pattern = "\\.csv$")) {
    cat(sprintf("  - %s\n", f))
  }
} else {
  df <- read_csv(CSV_PATH, show_col_types = FALSE)
  cat(sprintf("✓ Loaded %d rows from %s\n", nrow(df), basename(CSV_PATH)))
  cat(sprintf("\nColumns: %s\n", paste(names(df), collapse = ", ")))
}
```

```{r preview}
cat("📋 Data Preview (first 5 rows):\n")
kable(head(df, 5))
```

---

# Step 3: Validate Data

```{r validate-columns}
REQUIRED_COLUMNS <- c("LocationID", "SpeciesID", "Latitude", "Longitude")
OPTIONAL_COLUMNS <- c("DBH_cm", "Height_m", "TreeStatusID", "FieldNotes", "ExternalID")

cat("📋 Column Validation:\n")
cat(strrep("=", 50), "\n")

missing_required <- c()

for (col in REQUIRED_COLUMNS) {
  if (col %in% names(df)) {
    null_count <- sum(is.na(df[[col]]))
    if (null_count > 0) {
      cat(sprintf("  ⚠️  %s: %d NULL values\n", col, null_count))
    } else {
      cat(sprintf("  ✓ %s: OK\n", col))
    }
  } else {
    cat(sprintf("  ❌ %s: MISSING (required)\n", col))
    missing_required <- c(missing_required, col)
  }
}

cat("\n  Optional columns:\n")
for (col in OPTIONAL_COLUMNS) {
  if (col %in% names(df)) {
    valid_count <- sum(!is.na(df[[col]]))
    cat(sprintf("  ✓ %s: %d/%d values\n", col, valid_count, nrow(df)))
  } else {
    cat(sprintf("  - %s: not present\n", col))
  }
}

if (length(missing_required) > 0) {
  cat(sprintf("\n❌ Cannot import: missing required columns: %s\n", 
              paste(missing_required, collapse = ", ")))
} else {
  cat("\n✓ All required columns present\n")
}
```

```{r validate-fks}
conn <- get_connection()

cat("🔗 Foreign Key Validation:\n")
cat(strrep("=", 50), "\n")

# Check LocationIDs
db_locations <- dbGetQuery(conn, "SELECT LocationID, LocationName FROM shared.Locations")
csv_locations <- unique(na.omit(df$LocationID))
invalid_locations <- setdiff(csv_locations, db_locations$LocationID)

if (length(invalid_locations) > 0) {
  cat(sprintf("  ❌ Invalid LocationIDs: %s\n", paste(invalid_locations, collapse = ", ")))
  cat(sprintf("     Valid options: %s\n", paste(db_locations$LocationID, collapse = ", ")))
} else {
  cat(sprintf("  ✓ LocationIDs: all valid (%d unique)\n", length(csv_locations)))
}

# Check SpeciesIDs
db_species <- dbGetQuery(conn, "SELECT SpeciesID, CommonName FROM shared.Species")
csv_species <- unique(na.omit(df$SpeciesID))
invalid_species <- setdiff(csv_species, db_species$SpeciesID)

if (length(invalid_species) > 0) {
  cat(sprintf("  ❌ Invalid SpeciesIDs: %s\n", paste(invalid_species, collapse = ", ")))
  cat("     Valid options:\n")
  for (i in seq_len(nrow(db_species))) {
    cat(sprintf("       %d: %s\n", db_species$SpeciesID[i], db_species$CommonName[i]))
  }
} else {
  cat(sprintf("  ✓ SpeciesIDs: all valid (%d unique)\n", length(csv_species)))
}

dbDisconnect(conn)

if (length(invalid_locations) > 0 || length(invalid_species) > 0) {
  cat("\n❌ Fix foreign key issues before importing\n")
} else {
  cat("\n✓ All foreign keys valid\n")
}
```

```{r validate-coords}
cat("📍 Coordinate Validation:\n")
cat(strrep("=", 50), "\n")

lat_range <- range(df$Latitude, na.rm = TRUE)
lon_range <- range(df$Longitude, na.rm = TRUE)

cat(sprintf("  Latitude:  %.6f to %.6f\n", lat_range[1], lat_range[2]))
cat(sprintf("  Longitude: %.6f to %.6f\n", lon_range[1], lon_range[2]))

# Check if in Black Forest region
if (lat_range[1] > 47 && lat_range[2] < 49 && lon_range[1] > 7 && lon_range[2] < 9) {
  cat("\n  ✓ Coordinates in expected range (Black Forest region)\n")
} else {
  cat("\n  ⚠️  Coordinates may be outside expected area\n")
  cat("     Expected: Lat 47-49°, Lon 7-9° (Black Forest)\n")
}
```

---

# Step 4: Preview Import

```{r build-import}
# Build import dataframe
import_df <- df %>%
  mutate(
    LocationID = as.integer(LocationID),
    SpeciesID = as.integer(SpeciesID),
    VariantTypeID = 1L,  # field_measurement
    TreeStatusID = coalesce(as.integer(TreeStatusID), 1L),
    Height_m = as.numeric(Height_m),
    FieldNotes = coalesce(FieldNotes, ""),
    Position_WKT = sprintf("SRID=4326;POINT(%f %f)", Longitude, Latitude)
  )

# Add ExternalID to FieldNotes if present
if ("ExternalID" %in% names(df)) {
  import_df <- import_df %>%
    mutate(FieldNotes = paste0(FieldNotes, " | ExternalID: ", ExternalID))
}

cat(sprintf("📋 Import Preview (%d rows):\n", nrow(import_df)))
cat(strrep("=", 60), "\n")

kable(head(import_df %>% 
             select(LocationID, SpeciesID, Height_m, FieldNotes, Position_WKT), 5))

cat(sprintf("\n📊 Summary:\n"))
cat(sprintf("  Rows to insert: %d\n", nrow(import_df)))
cat(sprintf("  With height data: %d\n", sum(!is.na(import_df$Height_m))))
cat(sprintf("  With position: %d\n", sum(!is.na(import_df$Position_WKT))))
```

---

# Step 5: Execute Import

```{r import-config}
# ============================================================================
# IMPORT CONFIGURATION
# ============================================================================

DRY_RUN <- TRUE  # Set to FALSE for actual import

# ============================================================================
```

```{r execute-import}
execute_import <- function(import_df, dry_run = TRUE) {
  stats <- list(attempted = 0, inserted = 0, failed = 0, errors = c())
  
  conn <- get_connection()
  
  cat(sprintf("\n%s - %s %d rows\n",
              ifelse(dry_run, "🔍 DRY RUN", "⚡ LIVE IMPORT"),
              ifelse(dry_run, "Testing", "Inserting"),
              nrow(import_df)))
  cat(strrep("=", 60), "\n")
  
  for (i in seq_len(nrow(import_df))) {
    row <- import_df[i, ]
    stats$attempted <- stats$attempted + 1
    
    query <- sprintf("
      INSERT INTO trees.Trees (
        LocationID, SpeciesID, VariantTypeID, TreeStatusID,
        Height_m, FieldNotes, Position, CreatedBy
      ) VALUES (
        %s, %s, %s, %s, %s, %s,
        ST_GeomFromEWKT('%s'),
        'import_trees_simple.Rmd'
      )
      RETURNING VariantID
    ",
      row$LocationID,
      ifelse(is.na(row$SpeciesID), "NULL", row$SpeciesID),
      row$VariantTypeID,
      row$TreeStatusID,
      ifelse(is.na(row$Height_m), "NULL", row$Height_m),
      ifelse(row$FieldNotes == "", "NULL", dbQuoteString(conn, row$FieldNotes)),
      row$Position_WKT
    )
    
    tryCatch({
      dbExecute(conn, query)
      stats$inserted <- stats$inserted + 1
    }, error = function(e) {
      stats$failed <<- stats$failed + 1
      if (length(stats$errors) < 5) {
        stats$errors <<- c(stats$errors, sprintf("Row %d: %s", i, substr(e$message, 1, 80)))
      }
    })
  }
  
  if (dry_run) {
    dbRollback(conn)
    cat("🔄 Rolled back (dry run)\n")
  } else {
    dbCommit(conn)
    cat("✅ Committed to database\n")
  }
  
  dbDisconnect(conn)
  return(stats)
}

# Execute import
stats <- execute_import(import_df, dry_run = DRY_RUN)

cat(sprintf("\n📊 Import Statistics:\n"))
cat(sprintf("  Attempted: %d\n", stats$attempted))
cat(sprintf("  Inserted:  %d\n", stats$inserted))
cat(sprintf("  Failed:    %d\n", stats$failed))

if (length(stats$errors) > 0) {
  cat("\n❌ Sample errors:\n")
  for (err in stats$errors) {
    cat(sprintf("  %s\n", err))
  }
}

if (DRY_RUN) {
  cat("\n", strrep("=", 60), "\n")
  cat("ℹ️  This was a DRY RUN. No data was inserted.\n")
  cat("   Set DRY_RUN <- FALSE above and re-run to import.\n")
  cat(strrep("=", 60), "\n")
}
```

---

# Step 6: Verify Import

```{r verify, eval=!DRY_RUN}
# Run this after DRY_RUN = FALSE
conn <- get_connection()

recent <- dbGetQuery(conn, "
  SELECT 
    t.VariantID,
    l.LocationName,
    s.CommonName as Species,
    t.Height_m,
    ST_Y(t.Position) as Latitude,
    ST_X(t.Position) as Longitude,
    t.CreatedAt,
    t.CreatedBy
  FROM trees.Trees t
  LEFT JOIN shared.Locations l ON t.LocationID = l.LocationID
  LEFT JOIN shared.Species s ON t.SpeciesID = s.SpeciesID
  WHERE t.CreatedBy = 'import_trees_simple.Rmd'
  ORDER BY t.CreatedAt DESC
  LIMIT 10
")

dbDisconnect(conn)

if (nrow(recent) > 0) {
  cat(sprintf("📋 Recently imported trees (%d shown):\n", nrow(recent)))
  kable(recent)
} else {
  cat("No trees found with CreatedBy = 'import_trees_simple.Rmd'\n")
  cat("Run import with DRY_RUN <- FALSE first.\n")
}
```

---

# Summary

**Import completed!** 

Next steps:
1. If DRY_RUN was TRUE, set `DRY_RUN <- FALSE` and re-run Step 5
2. Verify imported data in Step 6
3. Check trees in database: `SELECT COUNT(*) FROM trees.Trees;`
