---
title: "🌳 Digital Forest Twin - Tree Data Importer (R)"
author: "Digital Forest Twin Project"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: bootstrap
    highlight: tango
---

# Overview

This R Markdown notebook provides a step-by-step workflow to import tree data from CSV files into the Digital Forest Twin database.

## Workflow

1. **Setup**: Load dependencies and connect to database
2. **Introspect**: Display available database schema
3. **Load CSV**: Load your data file
4. **Reference Data**: Explore species, locations, and sensor types
5. **Define Mapping**: Create column mappings
6. **Coordinate Mapping**: Handle lat/lon or x/y coordinates
7. **Preview**: Review data before insertion
8. **Insert**: Load data into database

---

# Step 1: Setup - Load Dependencies

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r dependencies}
# Load required packages
library(RPostgres)
library(DBI)
library(readr)
library(dplyr)
library(jsonlite)
library(knitr)

# Load environment variables
readRenviron("../../docker/.env")

cat("✓ Dependencies loaded\n")
```

---

# Step 2: Connect to Database

```{r database-config}
# Read connection parameters from environment
postgres_host <- Sys.getenv("POSTGRES_HOST", "localhost")
postgres_user <- Sys.getenv("POSTGRES_USER", "postgres")
postgres_password <- Sys.getenv("POSTGRES_PASSWORD")
postgres_db <- Sys.getenv("POSTGRES_DB", "postgres")
postgres_port <- as.integer(Sys.getenv("POSTGRES_PORT", "5432"))

if (postgres_password == "") {
  cat("❌ POSTGRES_PASSWORD not found in docker/.env\n")
} else {
  cat(sprintf("✓ Database config: %s@%s:%d/%s\n",
              postgres_user, postgres_host, postgres_port, postgres_db))
}
```

```{r database-connect}
# Create database connection
tryCatch({
  conn <- dbConnect(
    RPostgres::Postgres(),
    host = postgres_host,
    user = postgres_user,
    password = postgres_password,
    dbname = postgres_db,
    port = postgres_port
  )
  cat("✓ Database connection successful\n")
}, error = function(e) {
  cat(sprintf("❌ Database connection failed: %s\n", e$message))
})
```

---

# Step 3: Introspect Database Schema

```{r schema-query}
# Query database schema
query <- "
  SELECT
    table_schema,
    table_name,
    column_name
  FROM information_schema.columns
  WHERE table_schema IN ('shared', 'trees', 'sensor', 'pointclouds', 'environments')
  ORDER BY table_schema, table_name, ordinal_position
"

schema_result <- dbGetQuery(conn, query)
head(schema_result, 20)
```

```{r schema-display}
# Build schema info structure
schema_info <- list()

for (i in seq_len(nrow(schema_result))) {
  schema <- schema_result$table_schema[i]
  table <- schema_result$table_name[i]
  column <- schema_result$column_name[i]

  if (!(schema %in% names(schema_info))) {
    schema_info[[schema]] <- list()
  }
  if (!(table %in% names(schema_info[[schema]]))) {
    schema_info[[schema]][[table]] <- c()
  }

  schema_info[[schema]][[table]] <- c(schema_info[[schema]][[table]], column)
}

# Display schema
cat("\n", strrep("=", 80), "\n")
cat("DATABASE SCHEMA - Available Tables & Columns\n")
cat(strrep("=", 80), "\n")

for (schema_name in names(schema_info)) {
  cat(sprintf("\n📦 Schema: %s\n", schema_name))

  tables <- schema_info[[schema_name]]
  for (table_name in names(tables)) {
    columns <- tables[[table_name]]
    cat(sprintf("  📋 %s (%d columns)\n", table_name, length(columns)))

    for (i in seq_along(columns)) {
      cat(sprintf("     %2d. %s\n", i, columns[i]))
    }
  }
}
```

---

# Step 4: Load Reference Data

```{r reference-data}
# Load Species
species <- dbGetQuery(conn,
  "SELECT SpeciesID, CommonName, ScientificName FROM shared.Species ORDER BY CommonName"
)

# Load Locations
locations <- dbGetQuery(conn,
  "SELECT LocationID, LocationName FROM shared.Locations ORDER BY LocationName"
)

# Try to load SensorTypes
sensor_types <- tryCatch({
  dbGetQuery(conn,
    "SELECT SensorTypeID, SensorTypeName FROM sensor.SensorTypes ORDER BY SensorTypeName"
  )
}, error = function(e) {
  data.frame()
})

cat("✓ Reference data loaded\n")
```

```{r reference-data-display}
cat("\n", strrep("=", 80), "\n")
cat("📚 REFERENCE DATA - Use for mapping CSV values to database IDs\n")
cat(strrep("=", 80), "\n\n")

cat("📚 Species:\n")
kable(species, format = "html")

cat("\n📚 Locations:\n")
kable(locations, format = "html")

if (nrow(sensor_types) > 0) {
  cat("\n📚 SensorTypes:\n")
  kable(sensor_types, format = "html")
}
```

---

# Step 5: Load CSV File

```{r load-csv}
# Specify your CSV file path
csv_path <- "../../data/mathisle_250904.csv"  # Change this to your CSV path

# Check if file exists
if (!file.exists(csv_path)) {
  cat(sprintf("❌ CSV file not found: %s\n", csv_path))
} else {
  df <- read_csv(csv_path, show_col_types = FALSE)

  cat(sprintf("✓ Loaded CSV: %s\n", basename(csv_path)))
  cat(sprintf("  Rows: %d\n", nrow(df)))
  cat(sprintf("  Columns: %s\n", paste(names(df), collapse = ", ")))
}
```

```{r csv-preview}
cat(sprintf("\nFirst 3 rows of %s:\n", basename(csv_path)))
kable(head(df, 3), format = "html")
```

---

# Step 6: Coordinate Mapping Guide

If your CSV has separate lat/lon or x/y columns, follow these strategies:

```{r coordinate-guide, results='asis'}
cat("""
## Mapping Strategies

### 1️⃣ COMBINED COLUMN MAPPING (Recommended)
Format: `lat_lon:EPSG_CODE` or `x_y:EPSG_CODE`

Examples:
- For lat/lon columns: `'coordinates'` maps to `lat_lon:EPSG:4326`
- For x/y columns: `'utm_coords'` maps to `x_y:EPSG:32632`

Benefits: Automatic combination + CRS transformation

Available column names:
- **Latitude**: latitude, lat, lat_col, latitude_col, gps_latitude
- **Longitude**: longitude, lon, lon_col, longitude_col, gps_longitude
- **X/Easting**: x, x_col, easting, easting_col, utm_x, x_32632
- **Y/Northing**: y, y_col, northing, northing_col, utm_y, y_32632

### 2️⃣ MANUAL GEOMETRY STRING
Map to: `shared.Locations.Position` (or `trees.Trees.Position`, etc.)
Create WKT format before import: `POINT(lon lat)`

### 3️⃣ SEPARATE COORDINATES (Not Recommended)
❌ Don't map lat/lon to separate database columns
✅ Combine them first using option 1 or 2

## CRS Support

- **EPSG:4326** (WGS84) - default, no transformation needed
- **EPSG:32632** (UTM Zone 32N) - automatically transforms to WGS84
- **Other EPSG codes** - specify in mapping format

Example: `x_y:EPSG:32633` (UTM Zone 33N)

## Common Coordinates in Your Data

- **Mathisle**: gps_latitude, gps_longitude (WGS84)
- **EcoSense**: x_32632, y_32632 (UTM Zone 32N)
"""
)
```

---

# Step 7: Define Column Mapping

Edit the mapping below to configure for your CSV:

```{r mapping-definition}
# Define your column mapping
# Format: csv_column = "schema.table.column" or "lat_lon:EPSG:4326" or "x_y:EPSG:32632" or "SKIP"

mapping <- list(
  # Example for Mathisle trees CSV:
  # species_short = "shared.Species.CommonName",
  # gps_latitude = "lat_lon:EPSG:4326",  # Auto-detects gps_longitude
  # gps_longitude = "SKIP",              # Skip since lat_lon mapping handles it
  # DBH = "trees.Trees.FieldNotes",
  # TreeID = "trees.Trees.FieldNotes",
  # height = "trees.Trees.Height_m"
)

cat("Mapping defined. Edit above to configure for your CSV.\n\n")
cat("CSV Columns:\n")
print(names(df))

cat("\nCurrent Mapping:\n")
if (length(mapping) > 0) {
  for (csv_col in names(mapping)) {
    cat(sprintf("  '%s' -> %s\n", csv_col, mapping[[csv_col]]))
  }
} else {
  cat("  (empty - edit the cell above)\n")
}
```

---

# Step 8: Inspect CSV Values (LOOKUP)

Use this cell to examine specific columns before deciding on mapping:

```{r lookup-column}
# Change this to inspect any column
column_to_inspect <- "species_short"

if (column_to_inspect %in% names(df)) {
  unique_values <- unique(df[[column_to_inspect]])
  cat(sprintf("Sample values from '%s' (%d unique):\n\n",
              column_to_inspect, length(unique_values)))

  # Show first 10 unique values
  for (val in head(unique_values, 10)) {
    cat(sprintf("  - %s\n", val))
  }

  if (length(unique_values) > 10) {
    cat(sprintf("  ... and %d more\n", length(unique_values) - 10))
  }
} else {
  cat(sprintf("Column '%s' not found in CSV\n", column_to_inspect))
  cat(sprintf("Available columns: %s\n", paste(names(df), collapse = ", ")))
}
```

---

# Step 9: Save Mapping as JSON

```{r save-mapping}
# Create mapping directory if it doesn't exist
mapping_dir <- dirname(csv_path)
mapping_filename <- sprintf("%s_mapping.json", tools::file_path_sans_ext(basename(csv_path)))
mapping_path <- file.path(mapping_dir, mapping_filename)

# Save mapping to JSON
write_json(mapping, mapping_path, pretty = TRUE)

cat(sprintf("✓ Mapping saved to: %s\n\n", mapping_path))

# Display saved mapping
cat("Saved mapping:\n")
cat(readLines(mapping_path), sep = "\n")
```

---

# Step 10: Process Coordinates

```{r coordinate-processing}
detect_coordinate_columns <- function(df, format_spec) {
  "Detect lat/lon or x/y columns by flexible naming"
  col_names <- tolower(names(df))

  if (grepl("^lat_lon", format_spec, ignore.case = TRUE)) {
    # Find latitude column
    lat_patterns <- c("latitude", "lat", "gps_latitude", "gps_lat")
    lat_col <- NULL
    for (pattern in lat_patterns) {
      idx <- which(col_names == pattern)
      if (length(idx) > 0) {
        lat_col <- names(df)[idx[1]]
        break
      }
    }

    # Find longitude column
    lon_patterns <- c("longitude", "lon", "gps_longitude", "gps_lon")
    lon_col <- NULL
    for (pattern in lon_patterns) {
      idx <- which(col_names == pattern)
      if (length(idx) > 0) {
        lon_col <- names(df)[idx[1]]
        break
      }
    }

    return(list(lat = lat_col, lon = lon_col))

  } else if (grepl("^x_y", format_spec, ignore.case = TRUE)) {
    # Find x/easting column
    x_patterns <- c("x", "easting", "utm_x", "x_32632", "x_32633")
    x_col <- NULL
    for (pattern in x_patterns) {
      idx <- which(col_names == pattern)
      if (length(idx) > 0) {
        x_col <- names(df)[idx[1]]
        break
      }
    }

    # Find y/northing column
    y_patterns <- c("y", "northing", "utm_y", "y_32632", "y_32633")
    y_col <- NULL
    for (pattern in y_patterns) {
      idx <- which(col_names == pattern)
      if (length(idx) > 0) {
        y_col <- names(df)[idx[1]]
        break
      }
    }

    return(list(x = x_col, y = y_col))
  }

  return(list())
}

# Process coordinate columns
process_coordinates <- function(df, mapping) {
  "Process coordinate columns and create Position geometry"

  for (csv_col in names(mapping)) {
    target <- mapping[[csv_col]]

    if (is.null(target) || !is.character(target) || !grepl(":", target)) {
      next
    }

    parts <- strsplit(target, ":")[[1]]
    coord_type <- parts[1]
    crs <- parts[2]

    if (coord_type == "lat_lon") {
      coords <- detect_coordinate_columns(df, target)

      if (is.null(coords$lat) || is.null(coords$lon)) {
        cat(sprintf("⚠️  Could not find lat/lon columns\n"))
        next
      }

      cat(sprintf("📍 Processing %s: lat=%s, lon=%s, CRS=%s\n",
                 csv_col, coords$lat, coords$lon, crs))

      # Create WKT POINT geometry
      df <- df %>%
        mutate(Position = ifelse(
          !is.na(!!sym(coords$lon)) & !is.na(!!sym(coords$lat)),
          sprintf("POINT(%f %f)", !!sym(coords$lon), !!sym(coords$lat)),
          NA_character_
        ))

      valid_count <- sum(!is.na(df$Position))
      cat(sprintf("✓ Created %d Position geometries\n", valid_count))

    } else if (coord_type == "x_y") {
      coords <- detect_coordinate_columns(df, target)

      if (is.null(coords$x) || is.null(coords$y)) {
        cat(sprintf("⚠️  Could not find x/y columns\n"))
        next
      }

      cat(sprintf("📍 Processing %s: x=%s, y=%s, CRS=%s\n",
                 csv_col, coords$x, coords$y, crs))

      # Create WKT POINT geometry (note: R transformation requires sf package)
      if (crs != "EPSG:4326") {
        cat(sprintf("ℹ️  Note: CRS transformation from %s to EPSG:4326 requires sf package\n", crs))
        cat("   Install with: install.packages('sf')\n")
      }

      df <- df %>%
        mutate(Position = ifelse(
          !is.na(!!sym(coords$x)) & !is.na(!!sym(coords$y)),
          sprintf("POINT(%f %f)", !!sym(coords$x), !!sym(coords$y)),
          NA_character_
        ))

      valid_count <- sum(!is.na(df$Position))
      cat(sprintf("✓ Created %d Position geometries\n", valid_count))
    }
  }

  return(df)
}

# Apply coordinate processing
df <- process_coordinates(df, mapping)
```

---

# Step 11: Organize Data by Table

```{r apply-mapping}
apply_mapping <- function(df, mapping) {
  "Organize data by target table"
  table_dfs <- list()

  for (csv_col in names(mapping)) {
    target <- mapping[[csv_col]]

    if (is.null(target) || grepl(":", target)) {
      next  # Skip special formats
    }

    # Parse schema.table.column
    parts <- strsplit(target, "\\.")[[1]]
    if (length(parts) != 3) {
      cat(sprintf("⚠️  Invalid mapping for '%s': %s\n", csv_col, target))
      next
    }

    schema <- parts[1]
    table <- parts[2]
    column <- parts[3]
    table_key <- sprintf("%s.%s", schema, table)

    if (!(table_key %in% names(table_dfs))) {
      table_dfs[[table_key]] <- tibble()
    }

    # Add column to table tibble
    table_dfs[[table_key]][[column]] <- df[[csv_col]]
  }

  # Add Position column if it exists
  if ("Position" %in% names(df)) {
    if ("trees.Trees" %in% names(table_dfs)) {
      table_dfs[["trees.Trees"]]$Position <- df$Position
    } else if ("sensor.Sensors" %in% names(table_dfs)) {
      table_dfs[["sensor.Sensors"]]$Position <- df$Position
    } else if (length(table_dfs) > 0) {
      first_table <- names(table_dfs)[1]
      table_dfs[[first_table]]$Position <- df$Position
    }
  }

  return(table_dfs)
}

# Apply mapping
table_dfs <- apply_mapping(df, mapping)
cat(sprintf("✓ Organized data into %d tables\n", length(table_dfs)))
```

---

# Step 12: Preview Data

```{r preview-data}
cat("\n", strrep("=", 80), "\n")
cat("DATA PREVIEW - How data will be inserted into each table\n")
cat(strrep("=", 80), "\n\n")

for (table_name in names(table_dfs)) {
  table_df <- table_dfs[[table_name]]
  cat(sprintf("📊 %s (%d rows, %d columns)\n",
             table_name, nrow(table_df), ncol(table_df)))
  cat(sprintf("   Columns: %s\n\n",
             paste(names(table_df), collapse = ", ")))
  cat("   First 2 rows:\n\n")
  print(kable(head(table_df, 2), format = "html"))
  cat("\n")
}
```

---

# Step 13: Insert Data (Optional)

Uncomment and run this section to insert prepared data into the database.

```{r insert-optional, eval=FALSE}
# Template for inserting data
insert_table_data <- function(conn, table_name, df) {
  "Insert tibble/dataframe into specified table"

  parts <- strsplit(table_name, "\\.")[[1]]
  schema <- parts[1]
  table <- parts[2]

  tryCatch({
    dbWriteTable(
      conn,
      c(schema, table),
      df,
      append = TRUE,
      overwrite = FALSE
    )
    cat(sprintf("✓ Inserted %d rows into %s\n", nrow(df), table_name))
  }, error = function(e) {
    cat(sprintf("❌ Failed to insert into %s: %s\n", table_name, e$message))
  })
}

# Insert all tables
for (table_name in names(table_dfs)) {
  table_df <- table_dfs[[table_name]]
  if (nrow(table_df) > 0) {
    insert_table_data(conn, table_name, table_df)
  }
}
```

---

# Summary

## ✓ What you've done:

1. Connected to the Digital Forest Twin database
2. Introspected available database schema
3. Loaded reference data (Species, Locations, SensorTypes)
4. Loaded your CSV file
5. Defined column mappings
6. Processed coordinates (lat/lon or x/y)
7. Organized data by table
8. Previewed data before insertion

## Next Steps:

1. Review the data preview above
2. Mapping is saved to: `r mapping_path`
3. Data is ready in the `table_dfs` list for insertion
4. Uncomment and run Step 13 to insert into database
5. Or use the data in any other way you need

## Tips:

- You can reuse saved mappings by loading the JSON file
- Use the LOOKUP feature (Step 8) to inspect column values
- Modify the mapping cell to change column assignments
- Knit document sections individually to test each step

---

# Cleanup

```{r cleanup}
# Close database connection
if (!is.null(conn)) {
  dbDisconnect(conn)
  cat("✓ Database connection closed\n")
}
```
